---
title: "Final project"
author: "Christopher Tran"
date: "2/15/2022"
output: html_document
---

The data found here https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE18606 
contains micro-array serum expression profiles of male or female patients with or without kawasaki disease treated with IVIG.
Patients were then classified based on if they responded to IVIG treatment or not.  
The code below downloads the raw data and fold change data comparing IVIG non-responders to control or IVIG responders to controls 
your goal in this assignment is to return a set of genes that optimally differentiates IVIG non-responders and control individuals.
hint: For this assignment, you will need to: 

a) Generate PCA plots of the raw data colored by classification
b) Find a method of measurement that will allow you to monitor clustering efficiency of the raw data 
c) Take serial cuts through the data and return the cut that optimizes the distance between the two classifications.


```{r, error=TRUE}
############################################################################################
#### This function downloads raw GEO data and compiles it together into one data frame. ####
############################################################################################
#DS: The data set GSE number to download.
#gpl: The GPL number of the data set to download.
#gsm: A character string containing numbers indicating the treatment conditions to download.
#namestr: a list housing the column names for each dataset to download.
#duprm: if "T", all duplicated rows will be removed. if "F" all duplicate rows will be retained
GEORNAseqcompileR <- function(DS, gpl, gsm, namestr, duprm = "T"){
  pb <- txtProgressBar(min = 0, max = length(DS), style = 3)
  
  for(a in 1:length(DS)){
    #### Obtain names of dataset columns ####
    #########################################
    n <- namestr[[a]]
     #### Download GEO data ####
    ###########################
    gset <- suppressMessages(getGEO(DS[a], GSEMatrix =TRUE, AnnotGPL=TRUE))
    if (length(gset) > 1) idx <- grep(gpl[a], attr(gset, "names")) else idx <- 1
    gset <- gset[[idx]]
    # make proper column names to match toptable 
    fvarLabels(gset) <- make.names(fvarLabels(gset))
    # group names for all samples
    gsms <- gsm[a]
    sml <- c()
    for (i in 1:nchar(gsms)) { sml[i] <- substr(gsms,i,i) }
    # log2 transform
    ex <- exprs(gset)
    qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
    LogC <- (qx[5] > 100) ||
      (qx[6]-qx[1] > 50 && qx[2] > 0) ||
      (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
    if (LogC) { ex[which(ex <= 0)] <- NaN
    exprs(gset) <- log2(ex) }
    # set up the data and proceed with analysis
    sml <- paste("G", sml, sep="")    # set group names
    fl <- as.factor(sml)
    gset$description <- fl
    design <- model.matrix(~ description + 0, gset)
    colnames(design) <- levels(fl)
    fit <- lmFit(gset, design)
    cont.matrix <- makeContrasts(G1-G0, levels=design)
    fit2 <- contrasts.fit(fit, cont.matrix)
    fit2 <- eBayes(fit2, 0.01)
    tT <- topTable(fit2, adjust="fdr", sort.by="B", number=25000000)
    
    #### annotate with gene names ####
    ##################################
    tT <- subset(tT, select=c("ID","Gene.symbol")) #, ,"adj.P.Val","P.Value","t","B","logFC"
    ex2 <- data.table(ex)
    
    #### Subset columns of interest ####
    ####################################
    ex2 <- ex2[,grepl("[0-9]", strsplit(as.character(gsm[a]), split = "")[[1]]), with = FALSE]
    ex2$ID <- rownames(ex)
    mer <- merge(tT, ex2, by = "ID")
    
    #### retain only data columns and gene names ####
    #################################################
    mer <-data.table(mer) %>% setnames((colnames(mer)[grepl("GSM", colnames(mer))]), n)
    mer <- mer[,2:ncol(mer)]
    if(a ==1){compdt <- mer}
    if(a > 1){compdt <- merge(compdt,mer, by = "Gene.symbol", allow.cartesian = TRUE)}
    #### remove duplicated entries and genes without a name ####
    compdt <- compdt[!Gene.symbol == ""]    
    if(duprm == "T"){compdt <- compdt[!duplicated(Gene.symbol),]}
    setTxtProgressBar(pb, a)
  }
  close(pb)
  return(compdt)
}


#########################################################################################################
#### This function downloads fold change data from GEO and compiles it together into one data frame. ####
#########################################################################################################
#DS: The data set GSE number to download.
#gpl: The GPL number of the data set to download.
#gsm: A character string containing numbers indicating the treatment conditions to download.
#namestr: a list housing the column names for each dataset to download.
#duprm: if "T", all duplicated rows will be removed. if "F" all duplicate rows will be retained

GEOFCcompileR <- function(DS, gpl, gsm, namestr, duprm = "T"){
  pb <- txtProgressBar(min = 0, max = length(DS), style = 3)
  
  #compdt <- data.table()
  for(a in 1:length(DS)){
    #### Obtain names of dataset columns ####
    #########################################
    n <- namestr[[a]]
    
    #### Download GEO data ####
    ###########################
    gset <- suppressMessages(getGEO(DS[a], GSEMatrix =TRUE, AnnotGPL=TRUE))
    if (length(gset) > 1) idx <- grep(gpl[a], attr(gset, "names")) else idx <- 1
    gset <- gset[[idx]]
    # make proper column names to match toptable 
    fvarLabels(gset) <- make.names(fvarLabels(gset))
    # group names for all samples
    gsms <- gsm[a]
    sml <- c()
    for (i in 1:nchar(gsms)) { sml[i] <- substr(gsms,i,i) }
    # eliminate samples marked as "X"
    sel <- which(sml != "X")
    sml <- sml[sel]
    gset <- gset[ ,sel]
    
    # log2 transform
    ex <- exprs(gset)
    qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
    LogC <- (qx[5] > 100) ||
      (qx[6]-qx[1] > 50 && qx[2] > 0) ||
      (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
    if (LogC) { ex[which(ex <= 0)] <- NaN
    exprs(gset) <- log2(ex) }
    # set up the data and proceed with analysis
    sml <- paste("G", sml, sep="")    # set group names
    fl <- as.factor(sml)
    gset$description <- fl
    design <- model.matrix(~ description + 0, gset)
    colnames(design) <- levels(fl)
    fit <- lmFit(gset, design)
    cont.matrix <- makeContrasts(G1-G0, levels=design)
    fit2 <- contrasts.fit(fit, cont.matrix)
    fit2 <- eBayes(fit2, 0.01)
    tT <- topTable(fit2, adjust="fdr", sort.by="B", number=25000000)
    
    #### annotate with gene names ####
    ##################################
    #	tT <- subset(tT, select=c("ID","Gene.symbol")) #, ,"adj.P.Val","P.Value","t","B","logFC"
    
    #### Subset columns of interest ####
    ####################################	
    #ex2 <- data.table(tT[,c("Gene.symbol", "logFC", "P.Value")])
    ex2 <- data.table(subset(tT, select=c("Gene.symbol", "logFC", "P.Value")))
    
    #### rename columns and merge data.table ####
    #############################################
    setnames(ex2, colnames(ex2), n)
    if(a == 1){compdt <- ex2
    if(duprm == "T"){compdt <- compdt[!duplicated(Gene.symbol),]}
    }else{
      compdt <- merge(compdt, ex2, by = "Gene.symbol")
      if(duprm == "T"){compdt <- compdt[!duplicated(Gene.symbol),]}
    }
    setTxtProgressBar(pb, a)
  }
  close(pb)
  return(compdt)
}
```

```{r,Error=TRUE}
#Additional code/plots

# Version info: R 3.2.3, Biobase 2.30.0, GEOquery 2.40.0, limma 3.26.8
################################################################
#   Data plots for selected GEO samples
# BiocManager::install("maptools")
library(GEOquery)
library(limma)
library(umap)

# load series and platform data from GEO

gset <- getGEO("GSE18606", GSEMatrix =TRUE, getGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL6480", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

ex <- exprs(gset)
# log2 transform
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
  ex <- log2(ex) }

# box-and-whisker plot
dev.new(width=3+ncol(gset)/6, height=5)
par(mar=c(7,4,2,1))
title <- paste ("GSE18606", "/", annotation(gset), sep ="")
boxplot(ex, boxwex=0.7, notch=T, main=title, outline=FALSE, las=2)
dev.off()

# expression value distribution plot
par(mar=c(4,4,2,1))
title <- paste ("GSE18606", "/", annotation(gset), " value distribution", sep ="")
plotDensities(ex, main=title, legend=F)

# mean-variance trend
ex <- na.omit(ex) # eliminate rows with NAs
plotSA(lmFit(ex), main="Mean variance trend, GSE18606")

# UMAP plot (multi-dimensional scaling)
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 15, random_state = 123)
plot(ump$layout, main="UMAP plot, nbrs=15", xlab="", ylab="", pch=20, cex=1.5)

library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)

```


```{r,Error=TRUE}
## Including Plots

# You can also embed plots, for example:

#######################
#### Download data ####
#######################
suppressPackageStartupMessages(library(GEOquery))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(scatterplot3d))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(vsn))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(cluster))
suppressPackageStartupMessages(library(airway))
suppressPackageStartupMessages(library(sjmisc))
suppressPackageStartupMessages(library(ggfortify))
suppressPackageStartupMessages(library(Rgraphviz))
suppressPackageStartupMessages(library(RColorBrewer))


measured <- GEORNAseqcompileR(DS="GSE18606", 
                              gpl="GPL6480", 
                              gsm="000000000111111221111121222223333334433334344444", 
                              namestr=list(
                                c("ctrl1", "ctrl2", "ctrl3", "ctrl4", "ctrl5", "ctrl6", "ctrl7", "ctrl8", "ctrl9",
                                  "responder1", "responder2", "responder3", "responder4", "responder5",
                                  "responder6", "non_responder1", "non_responder2", "responder7", "responder8", 
                                  "responder9", "responder10", "responder11", "non_responder3", "responder12",
                                  "non_responder4","non_responder5","non_responder6","non_responder7","non_responder8",
                                  "responder_IVIG1","responder_IVIG2","responder_IVIG3","responder_IVIG4","responder_IVIG5","responder_IVIG6",
                                  "non_responder_IVIG1", "non_responder_IVIG2",
                                  "responder_IVIG7","responder_IVIG8","responder_IVIG9","responder_IVIG10",
                                  "non_responder_IVIG3",
                                  "responder_IVIG11",
                                  "non_responder_IVIG4","non_responder_IVIG5","non_responder_IVIG6","non_responder_IVIG7","non_responder_IVIG8")), 
                              duprm = "T")

ds <- c("GSE18606", "GSE18606")
gp <- c("GPL6480", "GPL6480")
gs <- c("00000000011111111111111111111XXXXXXXXXXXXXXXXXXX", "000000000XXXXXXXXXXXXXXXXXXXX1111111111111111111")
names <- list(c("Gene.symbol", "logFC", "P.Value"),c("Gene.symbol", "logFC_IVIG", "P.Value_IVIG"))
dtFC <- GEOFCcompileR(DS=ds, gpl=gp, gsm=gs, namestr=names, duprm = "T")

deg2 <- merge(dtFC, measured, by="Gene.symbol")
# additional informmation
my_id <- "GSE18606"
gse <- getGEO(my_id)
              
ex.gse<-exprs(gse[[1]])
f.gse<-fData(gse[[1]])
p.gse<-pData(gse[[1]])

summary(ex.gse)
lex.gse<-log2(ex.gse)

# additional box plot
boxplot(lex.gse,outline=FALSE)
boxplot(na.omit(lex.gse),outline=FALSE)

boxplot(na.omit(deg2[,2:53]),outline=FALSE)
boxplot(na.omit(log2(deg2[,2:53])),outline=FALSE)

summary(deg2)
boxplot(deg2[,2:53])
boxplot(na.omit(deg2[,2:53]))
boxplot(na.omit(log2(deg2[,2:53])),outline=FALSE)



ex=ex[apply(ex,MARGIN=1, function(x) { any(x!=0) }),]
ex.rpm = sweep(ex,MARGIN=2,colSums(ex)/1000000,FUN="/")
oldpar=par(mfrow=c(1,2),las=3,cex.axis=0.8)
boxplot(log2(ex+1),col="lightgreen")
boxplot(log2(ex.rpm+1),col="lightgreen")
par(oldpar)
# heatmap

cor.exgse<-cor(na.omit(ex.gse),use="c")
pheatmap(cor.exgse)

# pca <- prcomp(t(ex.gse))

```

```{r,Error=TRUE}
# a Generate PCA plots of the raw data colored by classification

# dataframe
deg2.1<-na.omit(deg2)
deg2.1<-deg2.1 %>% select(sort(names(deg2.1)))


control<-deg2.1[,1:9]
logFC<-deg2.1[,11:12]
p.value<-deg2.1[,29:30]

nonresponder.IVIG<-deg2.1[,13:20]
nonresponder<-deg2.1[,21:28]
gene.symbol<-deg2.1[,10]
responder.IVIG<-deg2.1[,31:41]
responder<-deg2.1[,42:53]


deg2.s<-as.data.frame(cbind(gene.symbol,p.value,logFC,control,nonresponder.IVIG,responder.IVIG,responder,nonresponder))
row.names(deg2.s)<- deg2.s[,1]

deg2.sr<-rotate_df(deg2.s, cn = T)

deg2.sr$response <- c(rep("N/A", 4),rep("control", 9),rep("resistant", 8),rep("responsive",23 ),rep("resistant", 8))
deg2.sr$response.type <- c(rep("N/A", 4),rep("control", 9),rep("acute", 8),rep("acute", 12),rep("convalescent", 11),rep("convalescent", 8))
deg2.values<-na.omit(deg2.sr[5:52,1:(length(deg2.sr)-2)])
deg2.response<-deg2.sr$response
deg2.class<-deg2.sr[5:52,(length(deg2.sr)-1):length(deg2.sr)]
deg2.stype<-paste(deg2.class[,1],"_",deg2.class[,2])
deg2.stype[1:9]<-"control"
deg2.sr$sample.type<-c(rep("N/A", 4),deg2.stype)
deg2.class$sample.type<-deg2.stype
deg2.ml<-na.omit(deg2.sr[5:52,])

# box plot
summary(deg2.1)

classes<-deg2.class[,1]
cols<-rainbow(length(unique(classes)))
point.cols<-cols[as.numeric( as.factor(classes))]


classes2<-deg2.class[,3]
cols2<-rainbow(length(unique(classes2)))
point.cols2<-cols2[as.numeric( as.factor(classes2))]


boxplot(t(deg2.values),outline=FALSE,col=point.cols)

boxplot(t(deg2.values),outline=FALSE,col=point.cols2)

oldpar=par(mfrow=c(3,1))
boxplot(control,outline=FALSE,col="red")
boxplot(nonresponder.IVIG,outline=FALSE,col="blue")
boxplot(responder.IVIG,outline=FALSE,col="green")
par(oldpar)


# PCA

pca.rpm=prcomp(deg2.values,retx=TRUE)

scatterplot3d(pca.rpm$x[,1:3],color=point.cols)

sc3=scatterplot3d(pca.rpm$x[,1:3],color=point.cols)
s3d.coords = sc3$xyz.convert(pca.rpm$x[,1:3])
text(s3d.coords$x,s3d.coords$y,labels=classes,cex=0.8,pos=4)
plot(pca.rpm)


scatterplot3d(pca.rpm$x[,1:3],color=point.cols2)
sc4=scatterplot3d(pca.rpm$x[,1:3],color=point.cols2)
s3d.coords2 = sc4$xyz.convert(pca.rpm$x[,1:3])
text(s3d.coords2$x,s3d.coords$y,labels=classes2,cex=0.8,pos=4)
plot(pca.rpm)

autoplot(pca.rpm, data = na.omit(deg2.sr[5:52,]), colour = 'response', label = TRUE, label.size = 3)

autoplot(pca.rpm, data = na.omit(deg2.sr[5:52,]), colour = 'sample.type', label = TRUE, label.size = 3)


vars=apply(deg2.values,2,var)

pca.2 = prcomp((deg2.values[,vars>1,]),retex=TRUE)
oldpar=par(mfrow=c(2,2),mar=c(4,4,1,1))
plot(pca.2$x[,1:2], pch=19, cex=1.5,col=point.cols)
legend(-15,3,col=cols,legend=unique(classes),cex=0.5,pch=19)
plot(pca.2$x[,c(1,3)], pch=19, cex=1.5,col=point.cols)
plot(pca.2$x[,2:3], pch=19, cex=1.5,col=point.cols)
scatterplot3d(pca.2$x[,1:3],color=point.cols,pch=19)
text(s3d.coords$x,s3d.coords$y,labels=classes,cex=0.8,pos=4)

oldpar=par(mfrow=c(2,2),mar=c(4,4,1,1))
plot(pca.2$x[,1:2], pch=19, cex=1.5,col=point.cols2)
legend(-15,3,col=cols,legend=unique(classes2),cex=0.5,pch=19)
plot(pca.2$x[,c(1,3)], pch=19, cex=1.5,col=point.cols2)
plot(pca.2$x[,2:3], pch=19, cex=1.5,col=point.cols2)
scatterplot3d(pca.2$x[,1:3],color=point.cols2,pch=19)
text(s3d.coords$x,s3d.coords$y,labels=classes2,cex=0.8,pos=4)




```


```{r, Error=TRUE}
# b Find a method of measurement that will allow you to monitor clustering efficiency of the raw data 

# heatmap

labels=rownames(data.matrix(deg2.values))
corr.dist=function(x) { as.dist(1-cor(t(x))) }

vars=apply(deg2.values,2,var)

select=order(vars,decreasing=T)[1:100]

redgreen <- function(n) {
 c(hsv(h=2/6, v=seq(1,0,length=n/2) ), hsv(h=0/6, v=seq(0,1,length=n/2)) )
}


heatmap.2(data.matrix(deg2.values[,select]),trace="none",
labCol=labels,labRow=F,col=redgreen(100),scale="row",
margins=c(7,7))


# clustering

d2<-dist(deg2.values)

oldpar=par(mfrow=c(1,1))
plot(hclust(d2,method="average"),label=rownames(deg2.values))
plot(hclust(d2,method="ward.D2"),label=rownames(deg2.values))
par(oldpar)



hc=hclust(corr.dist(deg2.values[,select]),method="ward.D2")
plot(hc,label=rownames(deg2.values))
rect.hclust(hc,h=3)
clusters=cutree(hc,h=3)



# k means


fit=kmeans(deg2.values,2)
fit$withinss
fit$withinss/fit$totss
fit$tot.withinss/fit$tots

fit2=kmeans(deg2.values,5)
fit$withinss
fit$withinss/fit$totss
fit$tot.withinss/fit$tots


fit.e=kmeans(t(deg2.values),2)
fit.e$withinss
fit.e$withinss/fit.e$totss
fit.e$tot.withinss/fit.e$tots

fit2.e=kmeans(t(deg2.values),5)
fit2.e$withinss
fit2.e$withinss/fit2.e$totss
fit2.e$tot.withinss/fit2.e$tots

```


```{r,Error=TRUE}

# c) Take serial cuts through the data and return the cut that optimizes the distance between the two classifications.


library(tximport)
library(DESeq2)
library(tidyverse)
library(limma)
library(breastCancerNKI)
suppressPackageStartupMessages(library(Biobase))
library(e1071)



# volcano plot
ggplot(data=deg2, aes(x=deg2$logFC, y=deg2$P.Value)) + geom_point()

# creating svm
ex.gse<-exprs(gse[[1]])

ge=ex.gse

e.m=factor(deg2.sr$sample.type[5:52])
levels(e.m)=c(0,1,2,3,4)

# pvalue
tt.pvals=p.value$P.Value

# dataframe
df=(as.data.frame(t(ge[order(tt.pvals),])))


# creating assess.prediction function
assess.prediction=function(truth,predicted) {
 predicted = predicted[ ! is.na(truth) ]
 truth = truth[ ! is.na(truth) ]
 truth = truth[ ! is.na(predicted) ]
 predicted = predicted[ ! is.na(predicted) ]
 cat("Total cases that are not NA: ",length(truth),"\n",sep="") 
 cat("Correct predictions (accuracy): ",sum(truth==predicted),
 "(",signif(sum(truth==predicted)*100/length(truth),3),"%)\n",sep="")

 TP = sum(truth==1 & predicted==1)
 TN = sum(truth==0 & predicted==0)
 FP = sum(truth==0 & predicted==1)
 FN = sum(truth==1 & predicted==0)
 P = TP+FN 
 N = FP+TN 
 cat("TPR (sensitivity)=TP/P: ", signif(100*TP/P,3),"%\n",sep="")
 cat("TNR (specificity)=TN/N: ", signif(100*TN/N,3),"%\n",sep="")
 cat("PPV (precision)=TP/(TP+FP): ", signif(100*TP/(TP+FP),3),"%\n",sep="")
 cat("FDR (false discovery)=1-PPV: ", signif(100*FP/(TP+FP),3),"%\n",sep="")
 cat("FPR =FP/N=1-TNR: ", signif(100*FP/N,3),"%\n",sep="")
}

# creating do.check function
do.check=function(data,model) {
 if ( is.null(data) ) { stop("New data must be specified") }
 if ( is.null(model) ) {
 stop("The model has not been trained yet!")
 }
}

# creating predict function
predict = function(M,newdata,...) {
 if ( inherits(M,"svm") && ! is.null(newdata) ) {
 has.na = apply(newdata,1,function(x) any(is.na(x)) )
 has.data = which(! has.na);
 has.na = which( has.na)
 pred = stats::predict(M,newdata[has.data,,drop=F],...)
 pred.with.na = pred[1] 
 pred.with.na[ has.data ] = pred
 pred.with.na[ has.na ] = NA
 pred.with.na 
 } else {
 stats::predict(M,newdata,...)
 }
}

# creating predictor.svm function
predictor.SVM = list(
 model = NULL,
 train = function(f,data,...) {
 predictor.SVM$model <<- svm(f, data,...) 
 },
 predict=function(newdata=NULL) {
 do.check(newdata,predictor.SVM$model)
 as.numeric( 
 as.character(predict(predictor.SVM$model,newdata)) > 0.5
 )
}
)

# creating cross validation  function
cross.validate=function(predictor,formula,data=NULL,
 method="random", N=1000, n.out=5,...) {
 if ( is.null(data) ) { stop("data must be specified") }
 f.str = deparse(formula)
 dependent.name = sub("\\s*~.*","",f.str)
 if ( ! dependent.name %in% names(data) ) {
 dependent.data = get(dependent.name,envir=environment(formula))
 data=cbind(dependent.data,data)
 names(data)[1] = dependent.name
 } else {
 ind = match(dependent.name,names(data))
 data = cbind( data[,ind,drop=F],data[,-ind,drop=F] )
 }
 truth = data[,dependent.name] 
 truth = truth[0] 
 prediction=numeric()
 
 for ( i in 1:N ) {
 leave.out = sample(nrow(data),size=n.out)
 training.data = data[-leave.out,,drop=F]
 test.data = data[leave.out,,drop=F]
 predictor$train(formula , data=training.data,...)
 pred=predictor$predict(test.data[,-1,drop=F])
 truth[ (length(truth)+1):(length(truth)+n.out) ] = 
 test.data[,dependent.name]
 prediction = c(prediction, pred)
 }
 list(truth=truth,prediction=prediction)
}

# svm
M=svm(e.m~.,df[,1:20,drop=F],kernel="linear")
assess.prediction(e.m,predict(M,df[,1:20]))


# running predictor svm 
predictor.SVM$train(e.m ~ . ,df[1:20])
assess.prediction(e.m, predictor.SVM$predict(df[1:20]))

# running cross validation
cv.SVM.20=cross.validate(predictor.SVM, e.m ~ . , df[1:20])
assess.prediction(cv.SVM.20$truth,cv.SVM.20$prediction)

cv.SVM.20r=cross.validate(predictor.SVM, e.m ~ . , df[1:20],kernel="radial")
assess.prediction(cv.SVM.20r$truth,cv.SVM.20r$prediction)

# associated genes
gene.logfc<-deg2[,c(1,2,4)]
gene.pvalue<-deg2[,c(1,3,5)]

deg.fc<-which(abs(gene.logfc[,3])> 1)
deg.p<-which(gene.pvalue[,3]<0.05)
gene.p<-deg.fc[deg.fc %in% deg.p]

f.deg=as.numeric()
for (x in gene.p) {
  f.deg<-gene.logfc[gene.p,1]
}
f.deg
```

